Bandit Multiclass Linear Classification: Efficient Algorithms for the Separable Case

### 2019-01-22 ICML-submission/from-weak-separability-to-strong-separability.tex
Changed p_i(y_t) to p_i(x_t) in (9) and (10)
Changed 0 to \zero; reformulated display in Thm 5
(Margin transformation theorem)

In this section, we consider case when the examples are weakly linearly
separable. Throughout this section, we assume without loss of generality that
all examples lie in the unit ball $\B(0,1) \subseteq \R^d$.\footnote{Instead of
working with feature vector $x_t$ we can work with normalized feature vectors
$\widehat{x}_t = \frac{x_t}{\norm{x_t}}$. It can be easily checked that if
$(x_1,y_1), (x_2,y_2), \dots, (x_T,y_T)$ are weakly linearly separable with
margin $\gamma$ and $\norm{x_t} \le R$ for all $t$, then the normalized examples
$(\widehat{x}_1,y_1), (\widehat{x}_2,y_2), \dots, (\widehat{x}_T,y_T)$ are weakly
linearly separable with margin $\gamma/R$.} Note that
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
alone does not guarantee a finite mistake bound in this setting, as weak linear
separability does not imply strong linear separability.

We construct a positive definite kernel function $k(\cdot, \cdot)$ whose
corresponding feature map $\phi(\cdot)$ transforms any sequence of \emph{weakly}
linearly separable examples to \emph{strongly} linearly separable sequence of
examples. Specifically, $\phi$ has the property that if a set of labeled
examples in $\B(0,1)$ is weakly linearly separable with a margin $\gamma$, then
after applying $\phi$ the examples become strongly linearly separable with a
margin $\gamma'$ and their norms are bounded by $2$. The parameter $\gamma'$ is
a function of the old margin $\gamma$ and the number of classes $K$, and is
specified in \autoref{theorem:margin-transformation} below.

The kernel function $k:\B(0,1) \times \B(0,1) \to \R$ is defined as
$$
k(x,x') = \frac{1}{1 - \frac{1}{2}\ip{x}{x'}_{\R^d}} \; .
$$
Note that kernel function $k(x,x')$ can be evaluated in $O(d)$ time.

Consider the classical real separable Hilbert space $\ell_2 = \{ x \in \R^\infty
~:~ \sum_{i=1}^\infty x_i^2 < + \infty \}$ equipped with the standard inner
product $\ip{x}{x'}_{\ell_2} = \sum_{i=1}^\infty x_i x'_i$. If we index the
coordinates of $\ell_2$ by $d$-tuples $(\alpha_1, \alpha_2, \dots, \alpha_d)$ of
non-negative integers, the feature map that corresponds to $k$ is $\phi: \B(0,1)
\to \ell_2$,
\begin{align}
& \left(\phi(x_1, x_2, \dots, x_d)\right)_{(\alpha_1, \alpha_2, \dots, \alpha_d)} = x_1^{\alpha_1} x_2^{\alpha_2} \dots x_d^{\alpha_d} \nonumber \\
& \qquad \cdot \sqrt{2^{-(\alpha_1 + \alpha_2 + \dots + \alpha_d)} \binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots, \alpha_d}}
\label{equation:phi}
\end{align}
where $\binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots,
\alpha_d} = \frac{(\alpha_1 + \alpha_2 + \dots + \alpha_d)!}{\alpha_1! \alpha_2!
\dots \alpha_d!}$ is the multinomial coefficient. It can be easily checked that
$$
k(x,x') = \ip{\phi(x)}{\phi(x')}_{\ell_2}.
$$
The last equality together with the formula for $k$ implies that $k(x,x) <
+\infty$ for any $x$ in $\B(0,1)$ and thus in particular implies that $\phi(x)$
indeed lies in $\ell_2$.

@@

\begin{theorem}[Margin transformation]
\label{theorem:margin-transformation}
Let $(x_1, y_1)$, $(x_2, y_2)$, $\dots$, $(x_T, y_T)$ in $\B(0,1) \times
\{1,2,\dots,K\}$ be a sequence of labeled examples that is weakly linearly
separable with margin $\gamma > 0$. Let $\phi$ be as defined in
equation~\eqref{equation:phi} and let
\begingroup
\allowdisplaybreaks
\begin{align*}
r & = 2 \left\lceil \frac{1}{4} \log_2(4K-3) \right\rceil + 1, \quad \quad s = \left \lceil \log_2(2/\gamma) \right \rceil, \\
\gamma_1 & = \frac{1}{2\sqrt{K}}  \\
& \ \cdot \left(376 \lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil \right)^{-\frac{1}{2} \lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil}, \\
\gamma_2 & = \frac{2^{s(s+1)r(K-1)} }{4\sqrt{K}(4K-5) 2^{K-1}} \\
& \ \cdot \left(2^{s+1} r(K-1) (4s+2)^2 \right)^{-(s+1/2)r(K-1)} \; .
\end{align*}
\endgroup
Then, the sequence of labeled examples transformed by $\phi$,
namely $(\phi(x_1), y_1), (\phi(x_2), y_2), \dots,
(\phi(x_T), y_T)$, is strongly linearly separable with margin $\gamma' =
\max\{\gamma_1, \gamma_2\}$. In addition, for all $t$ in $\cbr{1,\ldots,T}$,
$k(x_t,x_t) \leq 2$.
\end{theorem}

@@

\begin{corollary}[Mistake upper bound]
\label{corollary:weakly-separable-examples-mistake-upper-bound}
Let $K$ be a positive integer and let $\gamma$ be a positive real number. If
$(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in \B(0,1) \times \{1,2,\dots,K\}$
is a sequence of weakly separable labeled examples with margin $\gamma > 0$,
then the expected number of mistakes made by Algorithm~\ref{algorithm:kernelized}
with kernel $k(x,x') = \frac{1}{1 - \frac{1}{2}\ip{x}{x'}_{\R^d}}$
is at most $2^{\widetilde{O}(\min(K \log^2 \frac{1}{\gamma},
\sqrt{\frac{1}{\gamma}} \log K))}$.
\end{corollary}

@@

The conditions \eqref{equation:poly-pos} and~\eqref{equation:poly-neg} are
equivalent to that
\begin{align}
\label{equation:poly-pos-i}
\forall t \in \cbr{1,2,\dots,T}, y_t = i \quad \Rightarrow \quad p_i(y_t) \ge \frac{\gamma'}{2} \; , \\
\label{equation:poly-neg-i}
\forall t \in \cbr{1,2,\dots,T}, y_t \neq i \quad \Rightarrow \quad p_i(y_t) \le -\frac{\gamma'}{2} \; .
\end{align}
hold for all $i \in \{1,2,\dots,K\}$. We can thus fix $i$ and focus on
construction of one particular polynomial $p_i$.

Since examples $(x_1,y_1), (x_2, y_2), \dots, (x_T,y_T)$ are weakly linearly separable,
all examples from class $i$ lie in
$$
R_i^+ = \!\!\!\!\! \!\!\!\!\! \bigcap_{j \in \cbr{1,2,\dots,K} \setminus \cbr{i}} \cbr{x \in \B(0,1) ~:~ \ip{w_i^* - w_j^*}{x} \ge \gamma},
$$
and all examples from the remaining classes lie in
$$
R_i^- = \!\!\!\!\! \!\!\!\!\! \bigcup_{j \in \cbr{1,2,\dots,K} \setminus \cbr{i}} \cbr{x \in \B(0,1) ~:~ \ip{w_i^* - w_j^*}{x} \le -\gamma}.
$$
Therefore, to satisfy conditions~\eqref{equation:poly-pos-i}
and~\eqref{equation:poly-neg-i}, it suffices to construct $p_i$ such that
\begin{align}
\label{eqn:r-plus}
x \in R_i^+ \qquad & \Longrightarrow \qquad p_i(x) \ge \frac {\gamma'} 2 \; , \\
\label{eqn:r-minus}
x \in R_i^- \qquad & \Longrightarrow \qquad p_i(x) \le - \frac {\gamma'} 2 \; .
\end{align}

According to the well known Stone-Weierstrass theorem~\citep[see
e.g.][Section~10.10]{Davidson-Donsig-2010}, on a compact set, multivariate
polynomials uniformly approximate any continuous function. Roughly speaking, the
conditions \eqref{eqn:r-plus} and \eqref{eqn:r-minus} mean that $p_i$
approximates on $\B(0,1)$ a scalar multiple of the indicator function of the
intersection of $K-1$ halfspaces $\bigcap_{j \in \cbr{1,2,\dots,K} \setminus
\cbr{i}} \cbr{x: \ip{w_i^* - w_j^*}{x} \geq 0}$ while within margin $\gamma$ along
the decision boundary, the polynomial is allowed to attain arbitrary values.
It is thus clear such a polynomial exists.

@@

\begin{theorem}[Polynomial approximation of intersection of halfspaces I]
\label{theorem:polynomial-approximation-1}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1},
\norm{v_2}, \dots, \norm{v_m} \le 1$. Let $\gamma \in (0,1)$. There exists a
multivariate polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $p(x) \ge 1/2$ \\ for all $\displaystyle x \in R^+ = \bigcap_{i=1}^m \left\{ x \in \B(0,1) ~:~  \ip{v_i}{x} \ge \gamma \right\}$,
\item $p(x) \le -1/2$ \\ for all $\displaystyle x \in R^- = \bigcup_{i=1}^m \left\{ x \in \B(0,1) ~:~  \ip{v_i}{x} \le - \gamma \right\}$,
\item $\displaystyle \deg(p) = \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil$,
\item $\displaystyle \norm{p} \le \left( 188 \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil \right)^{\frac{1}{2} \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil}$.
\end{enumerate}
\end{theorem}

\begin{theorem}[Polynomial approximation of intersection of halfspaces II]
\label{theorem:polynomial-approximation-2}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1},
\norm{v_2}, \dots, \norm{v_m} \le 1$. Let $\gamma \in (0,1)$.
Define
$$
r = 2 \left\lceil \frac{1}{4} \log_2(4m + 1) \right\rceil + 1 \quad \text{and} \quad s = \left \lceil \log_2(1/\gamma) \right \rceil \; .
$$
Then, there exists a multivariate polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $\displaystyle p(x) \ge 1/2$ \\
for all $\displaystyle x \in R^+ = \bigcap_{i=1}^m \left\{ x \in \B(0,1) ~:~ \ip{v_i}{x} \ge \gamma \right\}$,

\item $\displaystyle p(x) \le - 1/2$ \\
for all $\displaystyle x \in R^- = \bigcup_{i=1}^m \left\{ x \in \B(0,1) ~:~ \ip{v_i}{x} \le - \gamma \right\}$,

>>>>

In this section, we consider case when the examples are weakly linearly
separable. Throughout this section, we assume without loss of generality that
all examples lie in the unit ball $\B(\zero,1) \subseteq \R^d$.\footnote{Instead of
working with feature vector $x_t$ we can work with normalized feature vectors
$\widehat{x}_t = \frac{x_t}{\norm{x_t}}$. It can be easily checked that if
$(x_1,y_1), (x_2,y_2), \dots, (x_T,y_T)$ are weakly linearly separable with
margin $\gamma$ and $\norm{x_t} \le R$ for all $t$, then the normalized examples
$(\widehat{x}_1,y_1), (\widehat{x}_2,y_2), \dots, (\widehat{x}_T,y_T)$ are weakly
linearly separable with margin $\gamma/R$.} Note that
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
alone does not guarantee a finite mistake bound in this setting, as weak linear
separability does not imply strong linear separability.

We construct a positive definite kernel function $k(\cdot, \cdot)$ whose
corresponding feature map $\phi(\cdot)$ transforms any sequence of \emph{weakly}
linearly separable examples to \emph{strongly} linearly separable sequence of
examples. Specifically, $\phi$ has the property that if a set of labeled
examples in $\B(\zero,1)$ is weakly linearly separable with a margin $\gamma$, then
after applying $\phi$ the examples become strongly linearly separable with a
margin $\gamma'$ and their norms are bounded by $2$. The parameter $\gamma'$ is
a function of the old margin $\gamma$ and the number of classes $K$, and is
specified in \autoref{theorem:margin-transformation} below.

The kernel function $k:\B(\zero,1) \times \B(\zero,1) \to \R$ is defined as
$$
k(x,x') = \frac{1}{1 - \frac{1}{2}\ip{x}{x'}_{\R^d}} \; .
$$
Note that kernel function $k(x,x')$ can be evaluated in $O(d)$ time.

Consider the classical real separable Hilbert space $\ell_2 = \{ x \in \R^\infty
~:~ \sum_{i=1}^\infty x_i^2 < + \infty \}$ equipped with the standard inner
product $\ip{x}{x'}_{\ell_2} = \sum_{i=1}^\infty x_i x'_i$. If we index the
coordinates of $\ell_2$ by $d$-tuples $(\alpha_1, \alpha_2, \dots, \alpha_d)$ of
non-negative integers, the feature map that corresponds to $k$ is $\phi: \B(\zero,1)
\to \ell_2$,
\begin{align}
& \left(\phi(x_1, x_2, \dots, x_d)\right)_{(\alpha_1, \alpha_2, \dots, \alpha_d)} = x_1^{\alpha_1} x_2^{\alpha_2} \dots x_d^{\alpha_d} \nonumber \\
& \qquad \cdot \sqrt{2^{-(\alpha_1 + \alpha_2 + \dots + \alpha_d)} \binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots, \alpha_d}}
\label{equation:phi}
\end{align}
where $\binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots,
\alpha_d} = \frac{(\alpha_1 + \alpha_2 + \dots + \alpha_d)!}{\alpha_1! \alpha_2!
\dots \alpha_d!}$ is the multinomial coefficient. It can be easily checked that
$$
k(x,x') = \ip{\phi(x)}{\phi(x')}_{\ell_2}.
$$
The last equality together with the formula for $k$ implies that $k(x,x) <
+\infty$ for any $x$ in $\B(\zero,1)$ and thus in particular implies that $\phi(x)$
indeed lies in $\ell_2$.

@@

\begin{theorem}[Margin transformation]
\label{theorem:margin-transformation}
Let $(x_1, y_1)$, $(x_2, y_2)$, $\dots$, $(x_T, y_T)$ in $\B(\zero,1) \times
\{1,2,\dots,K\}$ be a sequence of labeled examples that is weakly linearly
separable with margin $\gamma > 0$. Let $\phi$ be as defined in
equation~\eqref{equation:phi} and let
%\begingroup
%\allowdisplaybreaks
%\begin{align*}
%r & = 2 \left\lceil \frac{1}{4} \log_2(4K-3) \right\rceil + 1, \quad \quad s = \left \lceil \log_2(2/\gamma) \right \rceil, \\
%\gamma_1 & = \frac{1}{2\sqrt{K}}  \\
%& \ \cdot \left(376 \lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil \right)^{-\frac{1}{2} \lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil}, \\
%\gamma_2 & = \frac{2^{s(s+1)r(K-1)} }{4\sqrt{K}(4K-5) 2^{K-1}} \\
%& \ \cdot \left(2^{s+1} r(K-1) (4s+2)^2 \right)^{-(s+1/2)r(K-1)} \; .
%\end{align*}
%\endgroup
\begingroup
\allowdisplaybreaks
\begin{align*}
%r & = 2 \left\lceil \frac{1}{4} \log_2(4K-3) \right\rceil + 1, \quad \quad s = \left \lceil \log_2(2/\gamma) \right \rceil, \\
\gamma_1 = & \frac{ 
  \left[
    376 \lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil 
  \right]^{
    \frac{-\lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{{2}/{\gamma}} \right \rceil}{2}
  }
}{2\sqrt{K}}, 
\\
\gamma_2 = & \frac{
   \left(2^{s+1} r(K-1) (4s+2)^2 \right)^{-(s+1/2)r(K-1)} 
}{4\sqrt{K}(4K-5) 2^{K-1}}
\\
&  \cdot 2^{s(s+1)r(K-1)}
\; ,
\end{align*}
\endgroup
where $r = 2 \left\lceil \frac{1}{4} \log_2(4K-3) \right\rceil + 1$ and $s = \left \lceil \log_2(2/\gamma) \right \rceil$.
Then, the sequence of labeled examples transformed by $\phi$,
namely $(\phi(x_1), y_1), (\phi(x_2), y_2), \dots,
(\phi(x_T), y_T)$, is strongly linearly separable with margin $\gamma' =
\max\{\gamma_1, \gamma_2\}$. In addition, for all $t$ in $\cbr{1,\ldots,T}$,
$k(x_t,x_t) \leq 2$.
\end{theorem}

@@

\begin{corollary}[Mistake upper bound]
\label{corollary:weakly-separable-examples-mistake-upper-bound}
Let $K$ be a positive integer and let $\gamma$ be a positive real number. If
$(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in \B(\zero,1) \times \{1,2,\dots,K\}$
is a sequence of weakly separable labeled examples with margin $\gamma > 0$,
then the expected number of mistakes made by Algorithm~\ref{algorithm:kernelized}
with kernel $k(x,x') = \frac{1}{1 - \frac{1}{2}\ip{x}{x'}_{\R^d}}$
is at most $2^{\widetilde{O}(\min(K \log^2 \frac{1}{\gamma},
\sqrt{\frac{1}{\gamma}} \log K))}$.
\end{corollary}

@@

The conditions \eqref{equation:poly-pos} and~\eqref{equation:poly-neg} are
equivalent to that
\begin{align}
\label{equation:poly-pos-i}
\forall t \in \cbr{1,2,\dots,T}, y_t = i \quad \Rightarrow \quad p_i(x_t) \ge \frac{\gamma'}{2} \; , \\
\label{equation:poly-neg-i}
\forall t \in \cbr{1,2,\dots,T}, y_t \neq i \quad \Rightarrow \quad p_i(x_t) \le -\frac{\gamma'}{2} \; .
\end{align}
hold for all $i \in \{1,2,\dots,K\}$. We can thus fix $i$ and focus on
construction of one particular polynomial $p_i$.

Since examples $(x_1,y_1), (x_2, y_2), \dots, (x_T,y_T)$ are weakly linearly separable,
all examples from class $i$ lie in
$$
R_i^+ = \!\!\!\!\! \!\!\!\!\! \bigcap_{j \in \cbr{1,2,\dots,K} \setminus \cbr{i}} \cbr{x \in \B(\zero,1) ~:~ \ip{w_i^* - w_j^*}{x} \ge \gamma},
$$
and all examples from the remaining classes lie in
$$
R_i^- = \!\!\!\!\! \!\!\!\!\! \bigcup_{j \in \cbr{1,2,\dots,K} \setminus \cbr{i}} \cbr{x \in \B(\zero,1) ~:~ \ip{w_i^* - w_j^*}{x} \le -\gamma}.
$$
Therefore, to satisfy conditions~\eqref{equation:poly-pos-i}
and~\eqref{equation:poly-neg-i}, it suffices to construct $p_i$ such that
\begin{align}
\label{eqn:r-plus}
x \in R_i^+ \qquad & \Longrightarrow \qquad p_i(x) \ge \frac {\gamma'} 2 \; , \\
\label{eqn:r-minus}
x \in R_i^- \qquad & \Longrightarrow \qquad p_i(x) \le - \frac {\gamma'} 2 \; .
\end{align}

According to the well known Stone-Weierstrass theorem~\citep[see
e.g.][Section~10.10]{Davidson-Donsig-2010}, on a compact set, multivariate
polynomials uniformly approximate any continuous function. Roughly speaking, the
conditions \eqref{eqn:r-plus} and \eqref{eqn:r-minus} mean that $p_i$
approximates on $\B(\zero,1)$ a scalar multiple of the indicator function of the
intersection of $K-1$ halfspaces $\bigcap_{j \in \cbr{1,2,\dots,K} \setminus
\cbr{i}} \cbr{x: \ip{w_i^* - w_j^*}{x} \geq 0}$ while within margin $\gamma$ along
the decision boundary, the polynomial is allowed to attain arbitrary values.
It is thus clear such a polynomial exists.

@@

\begin{theorem}[Polynomial approximation of intersection of halfspaces I]
\label{theorem:polynomial-approximation-1}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1},
\norm{v_2}, \dots, \norm{v_m} \le 1$. Let $\gamma \in (0,1)$. There exists a
multivariate polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $p(x) \ge 1/2$ \\ for all $\displaystyle x \in R^+ = \bigcap_{i=1}^m \left\{ x \in \B(\zero,1) ~:~  \ip{v_i}{x} \ge \gamma \right\}$,
\item $p(x) \le -1/2$ \\ for all $\displaystyle x \in R^- = \bigcup_{i=1}^m \left\{ x \in \B(\zero,1) ~:~  \ip{v_i}{x} \le - \gamma \right\}$,
\item $\displaystyle \deg(p) = \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil$,
\item $\displaystyle \norm{p} \le \left( 188 \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil \right)^{\frac{1}{2} \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil}$.
\end{enumerate}
\end{theorem}

\begin{theorem}[Polynomial approximation of intersection of halfspaces II]
\label{theorem:polynomial-approximation-2}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1},
\norm{v_2}, \dots, \norm{v_m} \le 1$. Let $\gamma \in (0,1)$.
Define
$$
r = 2 \left\lceil \frac{1}{4} \log_2(4m + 1) \right\rceil + 1 \quad \text{and} \quad s = \left \lceil \log_2(1/\gamma) \right \rceil \; .
$$
Then, there exists a multivariate polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $\displaystyle p(x) \ge 1/2$ \\
for all $\displaystyle x \in R^+ = \bigcap_{i=1}^m \left\{ x \in \B(\zero,1) ~:~ \ip{v_i}{x} \ge \gamma \right\}$,

\item $\displaystyle p(x) \le - 1/2$ \\
for all $\displaystyle x \in R^- = \bigcup_{i=1}^m \left\{ x \in \B(\zero,1) ~:~ \ip{v_i}{x} \le - \gamma \right\}$,
